---
title: "Workshop 11 - Neighbourhood"
author: Sylvain SCHMITT
date: '`r Sys.Date()`'
output: 
  revealjs::revealjs_presentation:
    theme: blood
    highlight: pygments
    center: true
    fig_caption: true
    self_contained: false
    reveal_plugins: ["chalkboard"]
    reveal_options:
      slideNumber: true
      previewLinks: true
  # pdf_document: default
  # html_document:
  #   theme: readable
  #   toc: true
  #   toc_float: yes
csl: /home/sylvain/Documents/Bibliography/csl/mee.csl
bibliography: /home/sylvain/Documents/Bibliography/library.bib
link-citations: yes
---

# Introduction

## Setup

```{r setup, message=FALSE, warning=FALSE}
rm(list = ls()) ; invisible(gc()) ; set.seed(42)
library(knitr) 
library(tidyverse)
library(rstan)
library(bayesplot)
library(sf)
library(leaflet)
library(broom.mixed)
theme_set(bayesplot::theme_default())
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = T)
opts_chunk$set(echo = T, eval = T, cache = T, fig.height = 5)
```

## Origin

> [Cristina Barber's talk StanCon20](https://www.youtube.com/watch?v=w3w2rsozjJc&list=PLCrWEzJgSUqzI3goQEAKkDsHg72inmqbe&index=22&t=2s)
> [Cristina Barber et al. case study](https://mc-stan.org/users/documentation/case-studies/plantInteractions.html)

# Data 

## Paracou P16-C1

```{r, eval=F}
trees <- src_sqlite(file.path("..", "..", "PhD", "data", 
                              "Paracou","trees", "Paracou.sqlite")) %>% 
  tbl("Paracou") %>% 
  filter(CensusYear == 2015) %>% 
  filter(Plot == 16, SubPlot == 1) %>%
  mutate(SpeciesLong = paste0(substr(Genus, 1, 1), ". ", Species)) %>% 
  mutate(DBH = CircCorr/pi) %>% 
  collect()
```

## Let's simulate !

```{r}
N <- 100
S <- 10
trees <- data.frame(
  id = 1:N,
  x = runif(N, min = 0, max = 10),
  y = runif(N, min = 0, max = 10),
  dbh = rnorm(N, mean = 10, sd =  1),
  Species = sample(paste0("Sp", 1:S), N, replace = T)
)
```

## Spatialize it

```{r, message=F}
trees <- trees %>% 
  st_as_sf(coords = c("x", "y"),
                 crs = '+proj=utm +zone=22 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0')
crs <- '+proj=longlat +datum=WGS84'
```

## Prepare map

```{r }
palSp <- colorFactor("viridis", trees$Species)
map <- leaflet(data = st_transform(trees, crs = crs)) %>%
  addCircles(color = ~palSp(Species),
             label = ~ Species, radius = ~ dbh/50) %>% 
  addLegend(pal = palSp, values = ~ Species)
```

## View

```{r, echo=F}
map
```

# Neighbourhood interactions

## Interactions

> Between plants

* Survival
* Growth
* Recruitment

> Closer interacts more than distant

## Densities - Individual

```{r }
ind <- sample_n(trees, 1)
ind.neigh <- st_buffer(ind, 5)
```

## Densities - Neighbours

```{r, echo=F}
leaflet() %>%
  addCircles(data = st_transform(trees, crs = crs), 
             color = ~palSp(Species),
             label = ~ Species, radius = ~ dbh/20) %>% 
  addPolygons(data = st_transform(ind.neigh, crs = crs),
              color = "red")
```

## Densities - Value

```{r, message=F, warning=F}
st_intersection(trees, ind.neigh) %>% 
  nrow()
```

## Densities

> Lack:

* size
* identity
* distance

> Individual neighbour interactions

## Interaction

* a global parameter per individual
* local parameter for each neighbour (size, distance, ...)

## Neighbourhood Crowding index (NCI)


$$NCI_i = \sum_j dbh_j^\gamma e^{-\delta.d_{i,j}^2}$$

$$NCI_i = \sum_j exp[{\gamma.log(DBH_j)-(\delta.d_{i,j}^2})]$$

# Growth

## Growth

$$growth_i \sim \mathcal N (\alpha + \beta_1 . DBH_i + \beta_2 .NCI_i, \sigma)$$

$$growth_i \sim \mathcal N (\alpha + \beta_1 . DBH_i + \beta_2 . \sum_j dbh_j^\gamma e^{-\delta.d_{i,j}^2}, \sigma)$$

## Let's simulate !

```{r }
alpha <- 0.8 # intercept
beta1 <- 0.2 # dbh effect
beta2 <- 0.035 # NCI effect
gamma <- 0.1 # neighbour size effect 
delta <- 0.05 # neighbour distance effect
sigma <- 0.01 # residual variation
D <- as.matrix(st_distance(trees))
units(D) <- NULL
S <- matrix(rep(trees$dbh, N), nrow = N, ncol = N, byrow = T)
trees$growth <- rnorm(N,
                      mean = alpha +
                        beta1 * trees$dbh +
                        beta2 * rowSums(exp(gamma*log(S)-(D^2*delta))),
                      sigma)
```

## Growth - Have a look

```{r, echo=F}
ggplot(trees, aes(growth)) +
  geom_histogram()
```

## Growth - Prepare map

```{r }
pal <- colorNumeric("inferno", trees$growth)
map <- leaflet(st_transform(trees, crs = crs)) %>%
  addCircles(color = ~pal(growth),
             label = ~paste(round(growth), "mm"), radius = ~ dbh/20) %>% 
  addLegend(pal = pal, values = ~ growth)
```

## Growth - Have a look

```{r , echo=F}
map
```

# Pairwise Matrices

## Example

```{r}
sample_n(trees, 4) %>% 
  st_distance() %>% 
  as.matrix()
```

## Larger example

```{r}
(D <- sample_n(trees, 7) %>% 
  st_distance() %>% 
  as.matrix())
```

## Sparse matrix

```{r}
D[D > units::as_units(5, "m")] <- 0
D
```

# Segment function

## Sparse matrix

```{r}
M <- matrix(c(3, 0, 1, 3, 0, 0, 0, 0, 5, 4, 0, 4, 2, 1, 1, 0), 
            nrow = 4, byrow = T)
M
```

## `nb_b`

```{r, echo=F}
M
```

> amount of non 0 value in each row

```{r}
nb_b <- c(3, 0, 3, 3)
```

## `X`

```{r, echo=F}
M
```

> non 0 value in each row

```{r}
X <- c(3, 1, 3, 5, 4, 3, 2, 1, 1) 
```

## `pos` 

```{r, echo=F}
M
X
```

> position of the first non 0 value in each row for the X vector

```{r}
pos <- c(1, 4, 7) # 
```

## `nb_b` in R

```{r, echo=F}
M
```

> amount of non 0 value in each row

```{r}
n_nb <- rowSums(M != 0)
```

## `X` in R

```{r, echo=F}
M
```

> non 0 value in each row

```{r}
X <- M[M > 0]
```

## `pos` in R

```{r, echo=F}
M
X
```

> position of the first non 0 value in each row for the X vector

```{r}
pos[1] <- 1
for(i in 2:nrow(M))
  pos[i] <- pos[i-1] + n_nb[i-1]
```

# Matrix model

## Matrices

```{r}
D <- as.matrix(st_distance(trees))
D[D > units::as_units(5, "m")] <- 0
units(D) <- NULL
S <- matrix(rep(trees$dbh, N), nrow = N, ncol = N, byrow = T)
S[D == 0] <- 0
```

## Data in stan

$$growth_i \sim \mathcal N (\alpha + \beta_1 . DBH_i + \beta_2 . \sum_j dbh_j^\gamma e^{-\delta.d_{i,j}^2}, \sigma)$$

```{stan output.var="m1", eval=F, cache=F}
data{  
  int N ;                       // number of individuals  
  vector<lower=0> [N] dbh ;     // diameter at breast height
  vector [N] growth ;           // growth
  vector [N] D[N] ;             // distance matrix
  vector [N] S[N] ;             // size matrix
}
```

## Parameters in stan

$$growth_i \sim \mathcal N (\alpha + \beta_1 . DBH_i + \beta_2 . \sum_j dbh_j^\gamma e^{-\delta.d_{i,j}^2}, \sigma)$$

```{stan output.var="m1", eval=F, cache=F}
parameters{
  real alpha ;            // Intercept
  real beta[2] ;          // DBH and NCI slopes
  real<lower=0> gamma ;   // neighbour diameter effect
  real<lower=0> delta ;   // neighbour distance effect
  real<lower=0> sigma ;   // residual variation
}
```

## Transformed parameters in stan

$$growth_i \sim \mathcal N (\alpha + \beta_1 . DBH_i + \beta_2 . \sum_j dbh_j^\gamma e^{-\delta.d_{i,j}^2}, \sigma)$$

```{stan output.var="m1", eval=F, cache=F}
transformed parameters{  
  vector[N] NCI ;  
  vector[N] smat[N] ;
  vector[N] dmat[N] ;
  for(i in 1:N){
    for(j in 1:N){
     smat[i,j] = S[i,j]^gamma ;  
     dmat[i,j] = D[i,j]*D[i,j]*delta ;
    }
  }
  for(n in 1:N)
      NCI[n] = sum(smat[n] ./ exp(dmat[n])) ; // vectorial division
}
```

## Model in stan

$$growth_i \sim \mathcal N (\alpha + \beta_1 . DBH_i + \beta_2 . \sum_j dbh_j^\gamma e^{-\delta.d_{i,j}^2}, \sigma)$$

```{stan output.var="m1", eval=F, cache=F}
model{
  alpha ~ normal(0, 1) ;
  beta ~ normal(0, 1) ;
  gamma ~ normal(0, 1) ; 
  delta ~ normal(0, 1) ; 
  sigma ~ exponential(1) ;
  growth ~ normal(alpha + 
                  beta[1] * dbh +
                  beta[2] * NCI,
                  sigma) ;
}
```

## All in stan

$$growth_i \sim \mathcal N (\alpha + \beta_1 . DBH_i + \beta_2 . \sum_j dbh_j^\gamma e^{-\delta.d_{i,j}^2}, \sigma)$$

```{stan output.var="m1", eval=F, cache=F}
data{  
  int N ;                       // number of individuals  
  vector<lower=0> [N] dbh ;     // diameter at breast height
  vector [N] growth ;           // growth
  vector [N] D[N] ;             // distance matrix
  vector [N] S[N] ;             // size matrix
}
parameters{
  real alpha ;            // Intercept
  real beta[2] ;          // DBH and NCI slopes
  real<lower=0> gamma ;   // neighbour diameter effect
  real<lower=0> delta ;   // neighbour distance effect
  real<lower=0> sigma ;   // residual variation
}
transformed parameters{  
  vector[N] NCI ;  
  vector[N] smat[N] ;
  vector[N] dmat[N] ;
  for(i in 1:N){
    for(j in 1:N){
     smat[i,j] = S[i,j]^gamma ;  
     dmat[i,j] = D[i,j]*D[i,j]*delta ;
    }
  }
  for(n in 1:N)
      NCI[n] = sum(smat[n] ./ exp(dmat[n])) ; // vectorial division
}
model{
  alpha ~ normal(0, 1) ;
  beta ~ normal(0, 1) ;
  gamma ~ normal(0, 1) ; 
  delta ~ normal(0, 1) ; 
  sigma ~ exponential(1) ;
  growth ~ normal(alpha + 
                  beta[1] * dbh +
                  beta[2] * NCI,
                  sigma) ;
}
```

## Data in R

```{r}
mdata1 <- list(
  N = nrow(trees),
  dbh = trees$dbh,
  growth = trees$growth,
  D = D,
  S = S
)
```

## Fit

```{r, eval=F}
m1 <- stan_model("Wk11_models/m1.stan")
fit1 <- sampling(m1, mdata1, chains = 2, iter = 500,
                 include = F, pars = c("NCI", "smat", "dmat"))
save(m1, fit1, file = "Wk11_save/m1.Rdata")
```

## Summary

```{r}
load("Wk11_save/m1.Rdata")
tidyMCMC(fit1, rhat = T, drop.pars = F, robust = T, ess = T) %>% 
  mutate(expected = c(alpha, beta1, beta2, delta, gamma, sigma, NA)) %>% 
  kable()
```

## Trace

```{r}
traces <- mcmc_trace(as.array(fit1, pars = c("beta", "gamma", "delta"))) +
  geom_hline(aes(yintercept = expected), col = "red",
             data = data.frame(parameter = c("beta[1]", "beta[2]",
                                             "gamma", "delta"),
                               expected = c(beta1, beta2, gamma, delta)))
```

## Trace

```{r, echo=F}
traces
```

# Segmented model

## Segments

```{r}
n_nb <- rowSums(D != 0)
d <- D[D > 0]
s <- S[S > 0]
pos <- rep(NA, N)
pos[1] <- 1
for(i in 2:N)
  pos[i] <- pos[i-1] + n_nb[i-1]
```

## Data in stan

$$growth_i \sim \mathcal N (\alpha + \beta_1 . DBH_i + \beta_2 . \sum_j dbh_j^\gamma e^{-\delta.d_{i,j}^2}, \sigma)$$

```{stan output.var="m1", eval=F, cache=F}
data{  
  int N ;                       // number of individuals  
  int M ;                       // number of distances
  vector<lower=0> [N] dbh ;     // diameter at breast height
  vector [N] growth ;           // growth
  vector [M] d ;                // distance matrix
  vector [M] s ;                // size matrix
  int n_nb[N] ;                 // non zero values
  int pos[N] ;                  // first non zero position
}
```

## Parameters in stan

$$growth_i \sim \mathcal N (\alpha + \beta_1 . DBH_i + \beta_2 . \sum_j dbh_j^\gamma e^{-\delta.d_{i,j}^2}, \sigma)$$

```{stan output.var="m1", eval=F, cache=F}
parameters{
  real alpha ;            // Intercept
  real beta[2] ;          // DBH and NCI slopes
  real<lower=0> gamma ;   // neighbour diameter effect
  real<lower=0> delta ;   // neighbour distance effect
  real<lower=0> sigma ;   // residual variation
}
```

## Transformed parameters in stan

$$growth_i \sim \mathcal N (\alpha + \beta_1 . DBH_i + \beta_2 . \sum_j dbh_j^\gamma e^{-\delta.d_{i,j}^2}, \sigma)$$

```{stan output.var="m1", eval=F, cache=F}
transformed parameters{  
  vector[N] NCI ;  
  vector[M] svec ;
  vector[M] dvec ;
  for(m in 1:M){
     svec[m] = s[m]^gamma ;  
     dvec[m] = d[m]*d[m]*delta ;
  }
  for(n in 1:N)
    NCI[n]=sum(exp(log(segment(svec, pos[n], n_nb[n])) -
                  (segment(dvec, pos[n], n_nb[n])))) ;
}
```

## Model in stan

$$growth_i \sim \mathcal N (\alpha + \beta_1 . DBH_i + \beta_2 . \sum_j dbh_j^\gamma e^{-\delta.d_{i,j}^2}, \sigma)$$

```{stan output.var="m1", eval=F, cache=F}
model{
  alpha ~ normal(0, 1) ;
  beta ~ normal(0, 1) ;
  gamma ~ normal(0, 1) ; 
  delta ~ normal(0, 1) ; 
  sigma ~ exponential(1) ;
  growth ~ normal(alpha + 
                  beta[1] * dbh +
                  beta[2] * NCI,
                  sigma) ;
}
```

## All in stan

$$growth_i \sim \mathcal N (\alpha + \beta_1 . DBH_i + \beta_2 . \sum_j dbh_j^\gamma e^{-\delta.d_{i,j}^2}, \sigma)$$

```{stan output.var="m1", eval=F, cache=F}
data{  
  int N ;                       // number of individuals  
  vector<lower=0> [N] dbh ;     // diameter at breast height
  vector [N] growth ;           // growth
  vector [N] D[N] ;             // distance matrix
  vector [N] S[N] ;             // size matrix
}
parameters{
  real alpha ;            // Intercept
  real beta[2] ;          // DBH and NCI slopes
  real<lower=0> gamma ;   // neighbour diameter effect
  real<lower=0> delta ;   // neighbour distance effect
  real<lower=0> sigma ;   // residual variation
}
transformed parameters{  
  vector[N] NCI ;  
  vector[N] smat[N] ;
  vector[N] dmat[N] ;
  for(i in 1:N){
    for(j in 1:N){
     smat[i,j] = S[i,j]^gamma ;  
     dmat[i,j] = D[i,j]*D[i,j]*delta ;
    }
  }
  for(n in 1:N)
      NCI[n] = sum(smat[n] ./ exp(dmat[n])) ; // vectorial division
}
model{
  alpha ~ normal(0, 1) ;
  beta ~ normal(0, 1) ;
  gamma ~ normal(0, 1) ; 
  delta ~ normal(0, 1) ; 
  sigma ~ exponential(1) ;
  growth ~ normal(alpha + 
                  beta[1] * dbh +
                  beta[2] * NCI,
                  sigma) ;
}
```

## Data in R

```{r}
mdata2 <- list(
  N = nrow(trees),
  M = length(d) ,
  dbh = trees$dbh,
  growth = trees$growth,
  d = d,
  s = s,
  n_nb = n_nb,
  pos = pos
)
```

## Fit

```{r, eval=F}
m2 <- stan_model("Wk11_models/m2.stan")
fit2 <- sampling(m2, mdata2, chains = 2, iter = 500,
                 include = F, pars = c("NCI", "svec", "dvec"))
save(m2, fit2, file = "Wk11_save/m2.Rdata")
```

## Summary

```{r}
load("Wk11_save/m2.Rdata")
tidyMCMC(fit2, rhat = T, drop.pars = F, robust = T, ess = T) %>%
  mutate(expected = c(alpha, beta1, beta2, delta, gamma, sigma, NA)) %>% 
  kable()
```

## Trace

```{r}
traces <- mcmc_trace(as.array(fit2, pars = c("beta", "gamma", "delta"))) +
  geom_hline(aes(yintercept = expected), col = "red",
             data = data.frame(parameter = c("beta[1]", "beta[2]",
                                             "gamma", "delta"),
                               expected = c(beta1, beta2, gamma, delta)))
```

## Trace

```{r, echo=F}
traces
```

## Comparison

```{r }
lapply(lapply(c(fit1, fit2), 
       get_elapsed_time), data.frame) %>% 
  bind_rows(.id = "model") %>%
  mutate(model = recode_factor(model,
                               "1" = "Matrix model",
                               "2" = "Segmented model")) %>% 
  mutate(total = warmup + sample) %>% 
  arrange(total) %>% 
  kable(caption = "Model speed comparison between matrix and segmented.")
```
